---
title: "SBC LTER analysis"
author: "Anna Ramji"
format: html
editor: visual
---

```{r}

```

## Introduction

In this Quarto Document, I'll be working with data from the Santa Barbara Coast LTER sites, analyzing data using statistical analysis methods covered in the Master of Environmental Data Science program's Environmental Data Science (EDS) 222: Statistics for Environmental Data Science course taught by Tamma Carleton (tcarleton\@ucsb.edu).

## Background

(Putting essential background information here)

(setting up the problem)

(metadata, column info, etc.)

### Dataset 1: Kelp Forest dynamics: Fish abundance

Reed, D. and R. Miller. 2023. SBC LTER: Reef: Kelp Forest Community Dynamics: Fish abundance ver 38. Environmental Data Initiative. https://doi.org/10.6073/pasta/92003d368dfe121f6709ad1507412f66 (Accessed 2023-11-22)

### Dataset 2: CA Spiny Lobster Data

Reed, D. and R. Miller. 2023. SBC LTER: Reef: Abundance, size and fishing effort for California Spiny Lobster (Panulirus interruptus), ongoing since 2012 ver 9. Environmental Data Initiative. https://doi.org/10.6073/pasta/3595322687af94cd532620ad9db94c77 (Accessed 2023-11-22).

### Dataset 3: Feeding relationships

Santa Barbara Coastal LTER, J.E. Byrnes, D.C. Reed, B.J. Cardinale, K.C. Cavanaugh, S.J. Holbrook, and R.J. Schmitt. 2022. SBC LTER: Reef: Feeding relationships for kelp forest species ver 4. Environmental Data Initiative. https://doi.org/10.6073/pasta/e82a082f4440d469c84e11a265266004 (Accessed 2023-11-22).





# Context

## California Spiny Lobsters

*Panulirus interruptus*


https://cadivingnews.com/the-lifecycle-of-spiny-lobsters/

(3.25 inches = 82.55 milimeters)


El Niño
Barnard, P., Hoover, D., Hubbard, D. et al. Extreme oceanographic forcing and coastal response due to the 2015–2016 El Niño. Nat Commun 8, 14365 (2017). https://doi.org/10.1038/ncomms14365


California Department of Fish and Wildlife, Marine Species Portal : https://marinespecies.wildlife.ca.gov/california-spiny-lobster/true/

California Department of Fish and Wildlife. 2019 California Spiny Lobster, *Panulirus interruptus*, Enhanced Status Report.

Contributors: Jenny Hofmeister, Carlos Mireles, Julia Coates, Travis Buck, Tom Mason, Doug Neilson (2019)

- Spiny lobsters mature at about 5 years of age, or 2.5-inch carapace length. They have a complex, 2-year reproductive cycle from mating to the settlement of juvenile lobsters. Fecundity increases with size, and females produce one brood of eggs per year.
- 2.5 inches = 63.5 mm

Paper commonly referenced to for size and age estimates at different life cycle stages of California Spiny Lobster:

Engle J. M. (1979). Ecology and growth of juvenile California spiny lobster, panulirus interruptus (Randall) (Los Angeles, California, United States: University of Southern California).



### SBC LTER  long term kelp forest study sites



### Importing libraries

```{r libraries, warning=FALSE, message=FALSE}
library(tidyverse)
library(readr)
library(gt)
library(tufte)
library(feasts)
library(here)
library(patchwork)
library(janitor)
library(tsibble)
library(forecast)
library(lubridate)

library(vegan) # for diversity index

```

# Reading in the Data


```{r lobster_data}
# using the package `here()`

# setting the file path for our lobster abundance csv
lobster_abundance_fp <- here("data", "lobsters", "knb-lter-sbc.77.9",
                             "Lobster_Abundance_All_Years_20230831.csv")

# read in lobster abundance data from filepath
lobster_abd <- read_csv(lobster_abundance_fp, na = c("-99999")) |> 
  clean_names() |> 
  mutate("day" = lubridate::day(date)) # adding a "day" column for possible analysis later on 

# preliminary exploration shows us that the primary month that data was collected in is August (month == 8)

# making a tidied subset, finding average count and size per year per site
lobster_subset <- lobster_abd |> 
  select(year, month, day, date, site, count, size_mm) |> 
  group_by(year, site) |> 
  summarize(avg_count = round(mean(count, na.rm = TRUE), 3), 
            avg_size = round(mean(size_mm, na.rm = TRUE), 3))

lobster_subset
```


```{r lobster_plots}
# plot average lobster abundance per site per year
lobster_abd_plot <- ggplot(data = lobster_subset) + 
  geom_point(aes(x = year, y = avg_count,
                 color = site)) +
  geom_line(aes(x = year, y = avg_count,
                color = site)) +
  geom_vline(xintercept = 2015,  linetype = "dotted") + # showing 2015 el nino
  facet_wrap(~site) +
  theme_bw() +
  theme(strip.background = element_rect(fill = "transparent")) +
  labs(color = "Site",
       x = "Year",
       y = "Average Count",
       title = "Lobster Abundance Averaged per Site per Year",
       subtitle = "SBC LTER sites") +
  theme(legend.position = "none")

# view plot
lobster_abd_plot

# plot average lobster size per site per year
lobster_size_plot <- lobster_subset |> ggplot() + 
  geom_point(aes(x = year, y = avg_size,
                 color = site)) +
  geom_line(aes(x = year, y = avg_size,
                 color = site)) +
  # showing lobster size of sexual maturity 
  geom_hline(yintercept = 63.5, linetype = "dashed", alpha = 0.5) + 
  geom_hline(yintercept = 82.55, linetype = "dashed", alpha = 0.5) + 
  geom_vline(xintercept = 2015,  linetype = "dotted") + # showing 2015 el nino
  facet_wrap(~site) +
  theme_bw() +
  theme(strip.background = element_rect(fill = "transparent")) +
  labs(
    #color = "Site",
       x = "Year",
       y = "Average Size (mm)",
       title = "Lobster Size Averaged per Site per Year",
       subtitle = "organized by SBC LTER site") +
  theme(legend.position = "none")


lobster_size_plot
  
```
Based on these plots and our early data exploration, I found it difficult to make any meaningful claims about lobster abundance over time, as it was so challenging to find a way to average counts per site over time. I tried doing total counts per site per year, total counts per site per day during August over time, and average count per site per year. I found that the average count per site per year made the most sense, but the "0" or non-sightings in each site were overwhelming. 

Moving on to lobster size (carapace size, in mm), taking the average size per site per year showed much more interesting trends, with a significant dip in size coinciding with the 2015/2016 El Niño, which was the strongest over a 145 year record (Barnard et. al., 2017). 

The lower dashed horizontal line indicates the average estimated size (and corresponding age) at which the California Spiny Lobster reaches sexual maturity: around 2.5 inches (63.5 mm), or 58mm and 5 years old for Males and 70 mm and roughly 7 years old for Females (California Department of Fish and Wildlife, 2019), (Lindberg 1955, in Engle 1979). 

The higher dashed horizontal line indicates the minimum legal catch size: a carapace measuring 3.25 inches or 82.55 mm, or around 8-13 years old (estimated age range of 7-11 for Males, 8-13 for Females) (California Department of Fish and Wildlife, 2019). 

This shows us that for the most part, on average, the Santa Barbara Coast LTER sites are somewhat of a haven for California Spiny lobsters in their first few of years of breeding. As this graph shows an *average* carapace size, and lobsters can vary significantly in size depending on their age, these research sites are likely home to a wide age and developmental stage range of lobsters.

The vertical dotted line indicates the first year of the 2015-2016 El Niño. At all sites, there is a notable drop in average carapace length from 2014 to 2015. 



```{r water-temp-data}
# setting water temperature file path
water_temp_fp <- here("data", "water-temp", "knb-lter-sbc.13.29", "Bottom_temp_all_years_20230724.csv")

# reading in water temp data
water_temp <- read_csv(water_temp_fp, na = c("-99999")) |> 
  clean_names() |> 
  rename("date" = date_local, "time" = time_local) |> # cleaning names
  select(-serial) |> 
  mutate("year" = lubridate::year(date), # adding columns to join by
         "month" = lubridate::month(date),  # and filter later on 
         "day" = lubridate::day(date))


# prepping water temp data to compare with lobster data

# finding the average water temperatures in August (month = 8) per year per site
# to use with Lobster Abundance data, since observations/data are only collected
# in August each year

august_temps <- water_temp |> 
  filter(month %in% c(8)) |> # select August temps
  group_by(year, site) |> 
  summarize(avg_temp = mean(temp_c, na.rm = TRUE)) # find avg temp


august_temp_trends <- ggplot(data = august_temps) +
  geom_point(aes(x = year, 
                 y = avg_temp,
                 color = site)) +
  geom_line(aes(x = year, 
                y = avg_temp,
                color = site)) +
  theme_bw() +
  facet_wrap(~site) +
  theme(strip.background = element_rect(fill = "transparent")) +
  labs(title = "Average Water Temperature in August over Time",
       subtitle = "Ocean bottom temperatures (C) at SBC LTER sites", 
       x = "Year",
       y = "Average Water Temperature (C)") +
  theme(legend.position = "none")


august_temp_trends


# make a lobster abundance-friendly water temperature subset 
## (narrow data before joining)

#temp_lobster_plots <- august_temp_trends

#head(water_temp)
#tail(water_temp)
```


```{r lobster_temp_join}
# let's join the lobster subset and our temperature trends data to investigate the relationship between average temperature and average lobster carapace size 

lobster_temps <- left_join(lobster_subset, # average size per site per year
                           august_temps)

# preliminary plot of the joined data
lobster_temps_plot <- ggplot(data = lobster_temps) + 
  geom_jitter(aes(x = avg_temp,
                  y = avg_size)) +
  theme_bw() +
  labs(title = "Average lobster size by average water temperature", 
       subtitle = "California Spiny Lobsters, temperature taken at kelp forest bottoms, (C)", 
       x = "Average Temperature (C)",
       y = "Average Carapace Length (mm)")

# view plot
lobster_temps_plot


# ungrouping by site to find average size and temperature per year across all sites
lobster_yrly_averages <- lobster_temps |> 
  select()

# run multiple linear regression on temp and size 
lobster_temp_reg <- lm(formula = avg_size ~ avg_temp + site, data = lobster_temps)
summary(lobster_temp_reg)

```
Interpretation of these coefficients:

- **Intercept:** at the Arroyo Quemado Reef (site AQUE), our "dummy variable", when the temperature at the kelp forest floor ("bottom temperature") is 0 degrees Celsius, the expected lobster carapace size is 94.1 mm (nonsensical, from the SBC LTER records of reef bottom temperatures, which have been recorded every 15 minutes, ongoing since 2000, the minimum recorded temperature of all time in our data set is `r(min(water_temp$temp_c))` degrees Celsius). 
- **Slope: `avg_temp`:** Holding `site` constant (comparing average lobster carapace lengths and average water temperatures only to those of the same SBC LTER study site), at the AQUE site, given a one unit change in average temperature (an increase of 1 degree Celsius), we expect the average carapace length of lobsters to change by `r cat(round(lobster_temp_reg$coefficients['avg_temp'], 2))` mm.

Interpretation of our `site` coefficients:

- `siteCARP`: For the SBC LTER Carpenteria Reef study site, given a 1 degree Celsius increase in temperature, we expect to see a  `r cat(round(lobster_temp_reg$coefficients['siteCARP'], 2))` mm increase in average lobster carapace size as compared to the AQUE site expected slope of -1.14 mm per 1 degree increase in Celsius, for a total expected change in average lobster carapace length of `r cat(round(lobster_temp_reg$coefficients['siteCARP'] + lobster_temp_reg$coefficients['avg_temp'], 2))` mm.   

(moving forward, I will only describe the overall total expected change (`site___` + `avg_temp`)):

- `siteIVEE`: For the SBC LTER Isla Vista Reef study site, given a 1 degree Celsius increase in temperature, we expect to see a 0.04 mm *increase* in average lobster carapace size. 
- `siteMOHK`: For the SBC LTER Mohawk Reef study site, given a 1 degree Celsius increase in temperature, we expect to see a 2.48 mm *decrease* in average lobster carapace size.
- `siteNAPL`: For the SBC LTER Naples Reef study site, given a 1 degree Celsius increase in temperature, we expect to see a 6.61 mm *increase* in average lobster carapace size.


```{r}
# sum(is.na(lobster_abd_temp$temp_c))

```


```{r lobster_joined_group}

lobster_summary <- lobster_abd_temp |> 
  group_by(date, site) |> 
  summarize(mean_temp = round(mean(temp_c, na.rm = TRUE), 3),
            count = n())
  
lobster_summary

```


```{r lobster_temp_reg}
lobster_regression <- lm(formula = temp_c ~ count, data = lobster_abd_temp)

summary(lobster_regression)

```

```{r}

lobster_abd_temp |> 
  ggplot(data = lobster_abd_temp) +
  geom_jitter(aes(x = temp_c, y = count,
                  color = site)) +
  facet_wrap(~site) 

    


```

```{r}
ggplot(data = lobster_summary) +
  geom_line(aes(x = date, y = mean_temp))

```


```{r}
acf(water_temp$temp_c, lag.max=12)
```



```{r}

#lobster_abd_temp |> forecast::autoplot()

# as_tsibble(lobster_abd_temp) %>%
#   model(
#     classical_decomposition(value, type = "additive")
#   ) %>%
#   components() %>%
#   autoplot() 
```



```{r fish-data}
# setting file path for fish abundance
fish_abd_fp <- here("data", "fish_abundance", "knb-lter-sbc.17.38", "Annual_fish_comb_20230814.csv")

fish_abd <- read_csv(fish_abd_fp, na = c("-99999")) |> 
  clean_names() |> 
  select(-year, -month) |> 
  filter(taxon_class %in% c("Actinopterygii"))

head(fish_abd)
tail(fish_abd)
```

```{r fish_temp}

fish_temp <- left_join(fish_abd, water_temp, by = c("date", "site")) |> 
  drop_na(temp_c)

head(fish_temp)
```

```{r}
fish_temp_summary <- fish_temp  |> 
  group_by(date, scientific_name, site) |> 
  summarize(mean_temp = (mean(temp_c, na.rm = TRUE)),
            count = n())
fish_temp_summary

```


```{r iucn-data}

iucn_fp <- here("data", "redlist_species_data_all_fish", "assessments.csv")

iucn_fish <- read_csv(iucn_fp) |> 
  clean_names() |> 
  select(scientific_name, redlist_category, population_trend)

head(iucn_fish)

iucn_fish_names <- list(unique(iucn_fish$scientific_name))
iucn_fish_names
# cross-selecting species 

fish_temp_iucn <- inner_join(fish_temp, iucn_fish) |> 
  select(-group, -mobility, -vis, -quad, -side) |> 
  mutate(year = lubridate::year(date)) |> 
  relocate(year, .before = site) 

head(fish_temp_iucn)

# nrow()

```


```{r}
fish_temp_summary |> 
  ggplot(aes(x = date,
             y = count,
             color=scientific_name)) +
  geom_point() +
  theme(legend.position = "none")
```


```{r plotting_function}

jitter_plot <- function(.data) {
  ggplot(data = .data, 
         aes(x = mean_temp,
             y = count)) +
    geom_jitter(alpha = 0.5) +
    geom_smooth(method='lm', formula= y~x, color="lightcoral", se=F, size=1.5) +
    labs(x = "Mean temperature (C)",
       y = "Count",
       title = "SBC LTER Average temperature (Celsius) over time") +
    theme_bw() 
}


```


```{r temp_time_plot}

temp_time_plot <- jitter_plot(.data = fish_temp_summary)

temp_time_plot
```





```{r}
fish_count_plot <- fish_temp_summary |> 
  ggplot(aes(x = mean_temp,
             y = count)) +
  geom_jitter() +
  geom_smooth(method='lm', formula= y~x, color="lightcoral", se=F, size=1.5) +
  theme_bw() +
  labs(x = "Mean temperature (C)",
       y = "Count",
       title = "SBC LTER Average temperature (Celsius) over time")


fish_count_plot


```


```{r}
hist(fish_temp_summary$mean_temp)
#hist(fish_temp_summary$count)

```

```{r fish_count_plot_facet}

fish_count_plot_facet <- fish_temp_summary |> 
  ggplot(aes(x = mean_temp,
             y = count)) +
  geom_jitter(alpha = 0.5) +
  geom_smooth(method='lm', formula= y~x, color="lightcoral", se=F, size=1.5) +
  facet_wrap(~site)
  theme_bw() +
  labs(x = "Mean temperature (C)",
       y = "Count",
       title = "SBC LTER Average temperature (Celsius) over time")


fish_count_plot_facet


```


```{r site_selections}

carp <- fish_temp_summary |> 
  filter(site %in% c("CARP"))

carp_plot <-  jitter_plot(.data = carp) +
  labs(subtitle = "Carpenteria Test Site (CARP)")
carp_plot


```




```{r}



# fish_temp_summary |> 
# ggplot(aes(x = mean_temp,
#              y = count)) +
#   geom_jitter(alpha = 0.5) +
#   geom_smooth(method='lm', formula= y~x, color="lightcoral", se=F, size=1.5) +
#   facet_wrap(~scientific_name)
#   theme_bw() +
#   labs(x = "Mean temperature (C)",
#        y = "Count",
#        title = "SBC LTER Average temperature (Celsius) over time")

```



```{r}



```




make diversity index, look at relationship between water temp and diversity


group by site, date

summarize with shannon index
https://rdrr.io/cran/vegan/man/diversity.html 



Null hypothesis: water temperature has zero correlation with fish diversity
Alternative hypothesis: water temperature is correlated with fish diversity


Shannon or Shannon–Weaver (or Shannon–Wiener) index is defined as $H' = - \sum_{i} log_{b} p_{i} $, where $p_{i}$ is the proportional abundance of species $i$ and $b$ is the base od the logarithm. It is most popular to use natural logarithms, but some argue for base $b = 2$ (which makese sense, but no real difference).

The diversity function can find the total (or gamma) diversity of pooled communities with argument groups. The average alpha diversity can be found as the mean of diversities by the same groups, and their difference or ratio is an estimate of beta diversity (see Examples). The pooling can be based either on the observed abundancies, or all communities can be equalized to unit total before pooling; see Jost (2007) for discussion. Functions adipart and multipart provide canned alternatives for estimating alpha, beta and gamma diversities in hierarchical settings.

None of these diversity indices is usable for empty sampling units without any species, but some of the indices can give a numeric value. Filtering out these cases is left for the user.



```{r diversity}
fish_diversity_df <- fish_temp_iucn |> 
  select(year, site, count, temp_c, scientific_name, common_name, redlist_category, population_trend) |> 
  group_by(year, site)

fish_diversity <- fish_diversity_df |> 
  summarize(index = vegan::diversity(count, index = "shannon"))

head(fish_diversity)


# get average temperature per site over time (per year)
```


```{r avg-temps}
avg_temps_fish <- fish_diversity_df |> 
  summarize(avg_temp = round(mean(temp_c, na.rm = TRUE), 3)) 

avg_temps_subset <- avg_temps_fish |> 
  filter(!site %in% c("AHND", "SCTW" ))


carp_temps <- avg_temps_fish |> 
  filter(site == "CARP")

napl_temps <- avg_temps_fish |> 
  filter(site == "NAPL")

ivee_temps <- avg_temps_fish |> 
  filter(site == "IVEE")

mohk_temps <- avg_temps_fish |> 
  filter(site == "MOHK")

aque_temps <- avg_temps_fish |> 
  filter(site == "AQUE")

golb_temps <- avg_temps_fish |> 
  filter(site == "GOLB")


agg_avg_temps_plot <- ggplot() +
  geom_line(data = avg_temps_fish,
            aes(x = year, y = avg_temp)) 
  
```


```{r avg-temps-plot}
avg_temp_subset_plot <- ggplot(data = avg_temps_subset) +
  geom_line(aes(x = year, y = avg_temp)) +
  facet_wrap(~site) +
  theme_bw() +
  theme(strip.background = element_rect(fill = "transparent")) +
  labs(x = "Average Temperature (C)",
       y = "Year", 
       title = "Annual Average Temperature Over Time",
       subtitle = "at nine SBC LTER sites")

avg_temp_subset_plot

```
- note - biggest/strongest El Nino that California has ever experienced, between 2015 and 2016. Rise in ocean temperatures. 



1.. site year index
2. site year temp -- average per site per year
3. site year iucn status

```{r iucn-status-time}

iucn_status <- fish_temp_iucn |> 
  group_by(year, site, redlist_category) |> 
#  pivot_wider(names_from = 'redlist_category', )
#|> 
  summarize(iucn_count = n()) 

iucn_status$redlist_category <- factor(iucn_status$redlist_category,
                                       levels = c("Critically Endangered",
                                                  "Vulnerable",
                                                  "Least Concern",
                                                  "Data Deficient"))

iucn_status |> ggplot() +
  geom_jitter(aes(x = year, y = iucn_count,
                  color = redlist_category),
              alpha = 0.8) +
  # geom_line(aes(x = year, y = iucn_count,
  #               color = redlist_category)) +
  geom_smooth(aes(x = year, y = iucn_count,
                  color = redlist_category),
              method = "lm", se = TRUE) +
  scale_color_manual(limits = c("Critically Endangered",
                                "Vulnerable",
                                "Least Concern",
                                "Data Deficient"),
                     values = c("Critically Endangered"="red",
                                "Vulnerable"="yellow",
                                "Least Concern"="cornflowerblue",
                                "Data Deficient"="grey")) +
  facet_wrap(~redlist_category) +
  theme_bw() +
  theme(strip.background = element_rect(fill = "transparent")) +
  labs(color = "IUCN Redist Status",
       x = "Year",
       y = "Count",
       title = "Fish Population Demographics Over Time",
       subtitle = "across all SBC LTER sites")


carp_iucn_trend <- iucn_status |> 
  filter(site == "CARP")

carp_iucn_trend |> ggplot() +
  geom_jitter(aes(x = year, y = iucn_count,
                  color = redlist_category),
              alpha = 0.8) +
    geom_smooth(aes(x = year, y = iucn_count,
                  color = redlist_category), method = "lm", se = TRUE) +

   # geom_line(aes(x = year, y = iucn_count,
   #              color = redlist_category)) +
  scale_color_manual(limits = c("Critically Endangered",
                                "Vulnerable",
                                "Least Concern",
                                "Data Deficient"),
                     values = c("Critically Endangered"="red",
                                "Vulnerable"="yellow",
                                "Least Concern"="cornflowerblue",
                                "Data Deficient"="grey")) +
  theme_bw() +
  facet_wrap(~redlist_category) +
  theme(strip.background = element_rect(fill = "transparent")) +
  labs(color = "IUCN Redist Status",
       x = "Year",
       y = "Count",
       title = "Fish Population Demographics Over Time",
       subtitle = "SBC LTER Carpenteria site")

#carp_iucn_trend

#summary(lm(formula = ))

#legend_order <- levels(with(iucn_status, reorder(Labels, Percent)))
# scale_fill_discrete(breaks=c('Critically Endangered', 'Vulnerable', 'Least Concern', 'Data Deficient'))
# values=c("Critically Endangered"="red", "Vulnerable"="yellow","Least Concern"="lightblue","Data Deficient"="grey"))

```




```{r demographics-over-time}

status_temp_trend <- inner_join(iucn_status, avg_temps_fish)
status_temp_trend

diversity_temp_trend <- inner_join(status_temp_trend, fish_diversity)
diversity_temp_trend

summary(lm(formula = index ~ avg_temp, data = diversity_temp_trend))

```





```{r food-web}

food_web <- read_csv("data/food_web/knb-lter-sbc.56.4/byrnes_foodweb_table1.csv", na = c("-99999")) |> 
  clean_names() 


```

